State of the Practice for Lattice Boltzmann Method Software

### 2022-03-03 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Completed revision of Section 11 (Threats to Validity)

We developed a methodology for evaluating the state of the practice of
research software \citep{SmithEtAl2021}. This methodology can be instantiated
for a specific domain of research software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step of the steps in
the methodology. 
% of the next sections provide an overview of the specifics of identifying
% (Section~\ref{identifysoftware}) and filtering (Section~\ref{filtersoftware})
% the candidate software, our quantitative measurements
% (Section~\ref{empiricalmeasures}), the AHP ranking process
% (Section~\ref{AHP}), developer interviews (Section~\ref{SecSurvey}), and
% interactions with one of the key members of the assessment team -- the Domain
% Expert (Section~\ref{sec_vet_software_list}).

@@

This section examines potential threats to the validity of this state of the
practice assessment. These can be categorized into methodology and data
collection issues. The goal of this assessment isn't to rank the software, but
to use the ranking exercise as a means to understand the state of the practice
of LBM software development.

The measures listed in our measurement template may not be broad enough to
accurately capture some qualities. For example, there are only two measures of
surface robustness. The measurement of robustness could be expanded, as it
currently only measures unexpected input. Other faults could be introduced, but
could require a large investment of time to develop, and might not be a fair
measure for all packages. Similarly, reusability is assessed along the number of
code files and LOC per file. While this measure is indicative of modularity, it
is possible that some packages have many files, with few LOC, but the files do
not contain source code that is easily reusable. The files may be poorly
formatted, or the source code may be vague and have ambiguous identifiers.
Furthermore, the measurement of understandability relies on 10 random source
code files. It is possible that the 10 files that were chosen to represent a
software package may not be a good representation of the understandability of
that package.

Regarding data collection, a risk to the validity of this assessment is missing
or incorrect data. Some software package data may not have been measured due to
technology issues like broken links. This issue arose with the measurement of
Palabos, which had a broken link to its user manual, as noted in
Section~\ref{Sec_CompareArtifacts}.

Some pertinent data may not have been specified in public artifacts, or may be
obscure within an artifact or web-page. The use of unit testing and continuous
integration was mentioned in the artifacts of only three (ESPResSo, Ludwig,
Musubi) packages. However, interviews suggested a more frequent use of both unit
testing and continuous integration in the development processes than what was
observed from the initial survey of the artifacts. For example, OpenLB, pyLBM,
and TCLB use such methods during development despite this not being explicitly
clear from an analysis of the material available online.

Furthermore, design documentation was measured to be a ``less common'' artifact
in this assessment, but it is probable that such documentation is part of all
LBM packages. After all, developing SCS is not a trivial endeavour. It is likely
that many packages have such documentation but did not make it public, and due
to this the measured data is not a true reflection of software package quality.

\wss{There are more threats to validity.  Brainstorm.  Watch for them while 
editing.  Look at Ao paper.}

Problems with the measurement of qualities:

\begin{itemize}
\item The assumption that more code files is an indicator of reusability
(Section~\ref{reusabilityresults}) may not be a valid assumption.
\item Understandability was measured using only 10 random files
(Section~\ref{Sec_SurfUnderstandability}).  This could be improved.
\end{itemize}

Overall ranking with equally weighted qualities isn't realistic
(Section~\ref{Sec_OverallQuality}).

Threat to validity - the gap between measures of everything else and Musubi
(Section~\ref{identifysoftware}), and then the gap between the manual measures
and the automated measures (Section~\ref{repmetrics}).

From Section~\ref{Sec_CompareTools}: ``Interviews with the developers revealed a
potentially more frequent use of both unit testing and continuous integration,
compared to what was observed from studying the repository artifacts.''

Interviews suggested a more frequent use of both unit testing and continuous
integration in the development processes than what was observed from the initial
survey. For example, OpenLB, pyLBM, and TCLB use such methods during
development, despite this not being explicitly clear from an analysis of the
material available online.  The correctness and verifiability of such packages
is not measured well using surface analysis.

Section~\ref{repmetrics} compares ranking by best practices to the community's
ranking.  Using stars to estimate the community's ranking is not necessarily
accurate.

Tool usage: Interviews with the developers revealed a potentially more frequent
use of both unit testing and continuous integration, compared to what was
observed from studying the repository artifacts.

Maintainability is assumed to be improved by a higher ratio of comments to
source code (Section~\ref{Sec_Maintainability}), but this is not necessarily
true.

a threat to validity is that we have implicitly assumed documentation is
valuable when measuring quality.

threat to validity is the person doing the measuring.  In the previous exercise
we did do an experiment to measure this and convince ourselves that it isn't
that significant a risk.

We assume that a large number of source files is associated with increased
reusability (Section~\ref{reusabilityresults}), but this is not necessarily the
case.

@@

We analyzed the state of the practice of software development in the Lattice
Boltzmann Methods software domain by quantitatively and qualitatively measuring,
and comparing, 24 software packages along quality attributes. The software
qualities that were assessed in this report are listed in
Section~\ref{softwarequalities}. A methodology for assessing the state of the
practice of software development in SCS domains was presented. Domain packages
were assessed to answer the software development related research questions
listed in Section~\ref{secIntro} to understand how software quality is impacted
by software development choices, including principles, processes, and tools.
Software developers were interviewed to identify development pain points, and to
identify how software quality is ensured. Quantitative data was used to rank the
software packages using the AHP. The ranking designations were compared with
rankings from the software development community, and we found that many of our
top 10 ranked packages are ranked highly by a domain expert. Recommendations for
improving software along quality metrics were made... The findings of this
report can be used to guide future development of SCS, specifically along
quality attributes, and to reduce software quality failures.

@@

>>>>

We developed a methodology for evaluating the state of the practice of
research software \citep{SmithEtAl2021}. This methodology can be instantiated
for a specific domain of research software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step of the steps in
the methodology.

@@

The threats to validity can be categorized into methodology and data collection
issues.  With respect to our methodology, the measurement template may not be
detailed enough to accurately capture some qualities. For example, there are
only two measures of surface robustness. The measurement of robustness could be
expanded, as it currently only measures unexpected input. Other intentional
faults could be introduced to assess robustness, but this would require a larger
investment of measurement time. Similarly, reusability
(Section~\ref{reusabilityresults}) is assessed via the number of code files and
LOC per file. While this measure is indicative of modularity, it is possible
that some packages have many files, with few LOC, but the files do not contain
source code that is easily reusable. The files may be poorly formatted, or the
source code may be vague and have ambiguous identifiers. Maintainability may
also be incorrectly estimated, since it is assumed to be improved by a higher
ratio of comments to source code (Section~\ref{Sec_Maintainability}), but this
is not necessarily true. Furthermore, the measurement of understandability
(Section~\ref{Sec_SurfUnderstandability}) relies on 10 random source code files.
By random chance, the 10 files that were chosen may not be a good representation
of the understandability of that package.

In our methodology one individual measures each package and the beginning and
ending of the measurement phase is separated by several months.  There is a
threat that the measurement may include unconscious biases, or the reviewers
judgement may drift over time.  In previous work \citep{SmithEtAl2016}, we
demonstrated that the measurement process is reasonably reproducible by having 5
products graded by a second reviewer. The ranking via this independent review
was almost identical to the original ranking. As long as each grader uses
consistent definitions, the relative comparisons in the AHP results will be
consistent between graders.

Regarding data collection, a risk to the validity of this assessment is missing
or incorrect data. Some data may not have been measured due to technology issues
like broken links. This issue arose with the measurement of Palabos, which had a
broken link to its user manual, as noted in Section~\ref{Sec_CompareArtifacts}.
Some pertinent data may not have been specified in public artifacts, or may be
obscure within an artifact or web-page. For instance, the use of unit testing
and continuous integration was mentioned in the artifacts of only three
(ESPResSo, Ludwig, Musubi) packages (Section~\ref{SecSurfCorrectAndVerifiab}).
However, interviews suggested a more frequent use of both unit testing and
continuous integration. For example, OpenLB, pyLBM, and TCLB use such methods
during development despite this not being explicit from an analysis of the
material available online.

Furthermore, design documentation was measured to be a ``less common'' artifact
in this assessment, but it is probable that such documentation is part of all
LBM packages. After all, developing research software is not a trivial
endeavour. It is likely that many packages have such documentation but did not
make it public, and due to this the measured data may not be a true reflection
of software package quality.

Another data collection concern is the gap between measurement of Musubi and the
measurement of all other packages (Section~\ref{identifysoftware}).
Unfortunately we were not aware of the existence of Musubi until after the
primary data collection period was completed. This gap has a secondary effect
because it caused a mismatch between the time of manual and automated data
collection (Section~\ref{repmetrics}).  With time and resource constraints we
could not redo the manual measurement, but we did update the automated measures
(like the number of files, LOC, etc.) when Musubi was added.

Our comparison between our ranking and the community's ranking
(Section~\ref{repmetrics}) may not be valid.  Using stars to estimate the
community's ranking is not necessarily accurate.  Moreover, our overall AHP
ranking makes the unrealistic assumption of equal weighting between qualities
(Section~\ref{Sec_OverallQuality}).

A final threat to validity is conceptual.  Our work assumes that following best
practices leads to improved software quality.  As the advice in software
engineering guidelines show, we are not alone in this assumption.  Ideally
software qualities should be measured directly and then the impact of best
practices on those qualities should be assessed.  However, researchers in this
domain always face the challenge that many qualities, like maintainability,
reusability and verifiability, are difficult to measure directly.  Therefore, we
follow the usual approach of assuming that following procedures and adhering to
standards yields a higher-quality product \citep[p.\ 112]{VanVliet2000}.

@@

We analyzed the state of the practice of software development in the Lattice
Boltzmann Methods software domain by quantitatively and qualitatively measuring,
and comparing, 24 software packages along quality attributes. The software
qualities that were assessed in this report are listed in
Section~\ref{softwarequalities}. A methodology for assessing the state of the
practice of software development in scientific domains was presented. Domain
packages were assessed to answer the software development related research
questions listed in Section~\ref{secIntro} to understand how software quality is
impacted by software development choices, including principles, processes, and
tools. Software developers were interviewed to identify development pain points,
and to identify how software quality is ensured. Quantitative data was used to
rank the software packages using the AHP. The ranking designations were compared
with rankings from the software development community, and we found that many of
our top 10 ranked packages are ranked highly by a domain expert. Recommendations
for improving software along quality metrics were made... The findings of this
report can be used to guide future development of research software,
specifically along quality attributes, and to reduce software quality failures.

@@

The goal of this assessment isn't to rank the software, but to use the ranking
exercise as a means to understand the state of the practice of LBM software
development.
